---
title: "Predicting Exercise Barbell Reps"
author: "Rhowell"
date: "2022-10-19"
output:
  html_document:
    keep_md: yes
    theme: "cosmo"
    toc: true
    toc_float: true
    toc_collapsed: true
  github_document: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

## Data Import

```{r, cache = T, message=FALSE, warning=FALSE}
# load libraries
library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)

# save URL as variable
train_url <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# download data
download.file(url = train_url, destfile = "train_df.csv")
download.file(url = test_url, destfile = "test_df.csv")

#Reading the file
train_df <- read.csv(file = "train_df.csv")
test_df <- read.csv(file = "test_df.csv")

rbind(dim(train_df),
  dim(test_df))
```

## Cleaning data

We will download and split the training dataset into training and validation.

```{r, cache=TRUE}
# only keep numeric columns that are not blank
train_cleaned <- train_df %>% 
    mutate(classe = as.factor(classe)) %>%
    select(-"X") %>%
    discard(is.character) %>% 
    select_if(~ !any(is.na(.)))

test_cleaned <- test_df %>% 
    select(-c("X", "problem_id")) %>%
    discard(is.character) %>% 
    select_if(~ !any(is.na(.)))

# split training into training and validation sets
set.seed(45454)
in_train <- createDataPartition(train_cleaned$classe, p=0.70, list=F)
train_split <- train_cleaned[in_train, ]
validation_data <- train_cleaned[-in_train, ]
```

## k-Nearest Neighbors

Now we have the cleaned data sets, we will use the first model as k Nearest Neighbor using 5-fold cross-validation.

```{r, cache=TRUE}
# normalize data with min-max
min_max_norm <- function(x) {
     (x - min(x)) / (max(x) - min(x))
}
 
train_norm <- as.data.frame(lapply(train_split[,-56], min_max_norm)) %>%
   cbind(train_split[56])
validation_norm <- as.data.frame(lapply(validation_data[,-56], min_max_norm)) %>%
   cbind(validation_data[,56])
 
# check for NAs
rbind(anyNA(train_norm),
   anyNA(validation_norm))

# train kNN
trctrl <- trainControl(method = "repeatedcv", number = 5, repeats = 3)
set.seed(12121)
knn_fit <- train(classe ~., data = train_norm, method = "knn",
    trControl=trctrl,
    preProcess = c("center"),
    tuneLength = 10)

```

Accuracy is `r knn_fit$results[1,2]` with k optimized at `r knn_fit$results[1,1]`.

Let's test on the validation set.

```{r, cache=TRUE}
predict_knn <- predict(knn_fit, validation_norm)
confusionMatrix(validation_data$classe, predict_knn)
accuracy_knn <- postResample(predict_knn, validation_data$classe)
knn_oose <- 1 - as.numeric(confusionMatrix(validation_data$classe, predict_knn)$overall[1])
```

Accuracy is `r round(accuracy_knn[1], 4) `, and estimated out of sample error is `r round(knn_oose, 4)`.

## Random Forest

After looking at kNN, we will use Random Forest, another classification algorithm, to see how it compares. We will use 5-fold cross validation.

```{r, cache=TRUE}
controlRf <- trainControl(method="cv", 5)
rf_fit <- train(classe ~ ., data=train_split, method="rf", trControl=controlRf, ntree=100)
rf_fit
```

Accuracy is `r rf_fit$results[2,2]` with trees optimized at `r rf_fit$results[2,1]`.


```{r, cache=TRUE}
predict_rf <- predict(rf_fit, validation_data)
confusionMatrix(validation_data$classe, predict_rf)
accuracy_rf <- postResample(predict_rf, validation_data$classe)
oose_rf <- 1 - as.numeric(confusionMatrix(validation_data$classe, predict_rf)$overall[1])
```

Accuracy is `r round(accuracy_rf[1], 4)`, and estimated out of sample error is `r oose_rf`.

## Predicting the test set

To predict the results of the test, we will use the Random Forest model.

```{r, cache=TRUE}
predict_test <- predict(rf_fit, test_cleaned)
rbind(predict_test)
```


